---
title: "DSKUS EDA"
output: html_document
date: '2022-04-21'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(corrplot)
library(reshape2)
library(randomForest)
```

```{r}
countryCodes = read_csv("countryCode_to_region.csv") %>%
  select(`alpha-3`,region)
```


```{r}
dat = read_excel("food_security.xlsx",sheet = 2)
agg_dat = dat %>%
  pivot_wider(names_from = var_name) %>%
  dplyr::select(!date) %>%
  dplyr::group_by(iso3c) %>%
  summarize(across(everything(),~mean(.x,na.rm=TRUE))) 
  #pivot_longer(!iso3c,names_to = "exp_var",values_to="values")
```

```{r}
corr_dat = agg_dat %>%
  select(!c(iso3c,fies_insecurity_pct)) %>%
  cor(use="complete.obs",method='spearman') -> M
corrplot(M,type="upper",method = 'number',order="FPC")

```


We wanted to understand at a country level if there were specific variables we can utilize to predict how food insecurity would change from the affects of climate change.While the response variable Global Food Security Index (GFSI) was collected from Economist Intelligence Unit (EIU) where they have rated how citizens of countries experience food insecurity, where higher scores indicate greater food insecurity. We notice that there are five variables that have a strong Spearman correlation coefficient with food security. They are the Logistics Performance Index, percent formally employed, cereal yield, Worldwide Governance Indicators (WGI), and Per-capita household final consumption expenditure (HFCE). Due to the high correlation it is worth investigating further.

```{r}
long_agg = agg_dat %>%
  select(iso3c,gfsi_2019,logistics_idx,pct_formal_employ,cereal_yield,wgi_avg,hfce) %>%
  pivot_longer(!c(iso3c,gfsi_2019),names_to='vars',values_to='values') %>%
  inner_join(countryCodes, by = c("iso3c"="alpha-3"))
```

```{r}
vars.labs <- c("Cereal Yield","HFCE","Logistics Index","Percent in Paid Employed","World Governance Index")
names(vars.labs) <- c("cereal_yield",'hfce','logistics_idx','pct_formal_employ','wgi_avg')
long_agg %>%
  filter( values < 2.307727e+12 ) %>%
ggplot(aes(x=values,y = gfsi_2019)) + 
  geom_point(aes(colour=region)) + 
  facet_wrap(~vars,scales="free_x", 
             labeller = labeller(vars=vars.labs)) + 
  geom_smooth(se=FALSE, colour='#949494',size=.9) + 
  ylab("Global Food Security Index") + 
  xlab("Variable Values") + 
  theme_bw() +
  ggtitle("Relationship of variables on Country's Food Security")
ggsave("food_security_pairs.jpg")
```


From the graphic above we can see that all the variables have strong relationships with the general food security. Particularly we see that as the number of individuals with consistent paid employment increases the higher the food security will be. This is not a novel or surprising conclusion as it is intuitively what we believe to be true. Conversely the HFCE is a measurement of consumer spending in a country and countries with the most spending are able to ensure food security as well. Both of these measurements are able to show a country's citizens general ability to generate income and consume goods. The data's results are not novel and we expected that richer countries would be able to provide greater food security. The value of these measurements is that we hope to create simulations to predict the change in food security if climate change would force an economic shutdown. We also see that the logistic index and cereal yield have a strong effect on food security as well. This is to be expected as the domestic production, as well as the ability to transport the food from production to consumer is vitally important.

Due to the large co-linearity of the data some form of information gain algorithm may be more successful in creating accurate predictions. For an initial baseline model using a random forest will enable us to get initial estimates for the relative importance of the variables.

```{r}
model_dat = agg_dat %>%
  select(c(gfsi_2019,hfce,cereal_yield,logistics_idx,pct_formal_employ,wgi_avg)) %>%
  drop_na()
rf.fit <- randomForest(gfsi_2019 ~ ., data=model_dat,ntree= 1000, keep.forest=FALSE,importance=TRUE)
```




```{r}
ImpData <- as.data.frame(importance(rf.fit))
ImpData$Var.Names <- c("HFCE","Cereal Yield","Logistics Index","Percent in Paid Employed","World Governance Index")




ggplot(ImpData, aes(x=reorder(Var.Names,`%IncMSE`), y=`%IncMSE`)) +
  geom_segment( aes(x=Var.Names, xend=Var.Names, y=0, yend=`%IncMSE`), color="skyblue") +
  geom_point(aes(size = IncNodePurity), color="blue", alpha=0.6) +
  theme_light() +
  coord_flip() +
  theme_bw() +
  xlab("Variable Names") +
  ggtitle("Variable Importance based on Permuation Measures")
ggsave("feature_importance_foodSecurity.jpg")
```

From the above graphic we see that the variables that contributed most to accuracy was the percent of citizens that had stable employment. This is consistent with what we anticipated It seems that although the HFCE and Cereal Yield had strong correlations it does not provide has much information gain as the logistics index and percent of paid employed. The other large indicator is the World Governance Index which is as well expected that governments with social programs are able to assist the citizens who face unemployment are unable to obtain food. From our initial findings we believe that obtaining more granular data on the logistics index of countries and world governance index measurements will enable us to understand in more detail how these two variables interact with food security. As well as enable us to create a more accurate simulation of the effect on these measurements.




